---
title: "NYC flight delay analysis"
author: "Cole"
date: "2026-03-01"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Objective

This analysis evaluates factors associated with arrival delays for flights departing New York City airports. The goal is to identify actionable booking decisions passengers can make to reduce the likelihood of significant delays. Data were obtained from the `nycflights13` dataset containing commercial flights departing NYC airports in 2013.

### Stakeholders

- Airline passengers planning travel
- Business travelers sensitive to arrival times
- Travel planners seeking lower delay risk options

```{r libraries, echo=FALSE}
library(tidyverse)
library(nycflights13)
library(arrow)
library(janitor)
```


```{r retrieve flight data, echo=FALSE}

write_csv(flights, "data/raw/flights.csv")
write_csv(airlines, "data/raw/airlines.csv")
write_csv(weather, "data/raw/weather.csv")
write_csv(airports, "data/raw/airports.csv")

```
### Data cleaning

Flight data were enriched by merging airline and airport metadata. Duplicate and incomplete flight records were removed, and missing delay values were excluded. Datetime features and categorical delay classifications were created to support comparative analysis.

```{r data cleaning reading and joining, echo=FALSE}

flights_raw <- read_csv("data/raw/flights.csv")

airlines_raw <- read_csv("data/raw/airlines.csv") |>
  rename(airline_name = name)

airports_raw <- read_csv("data/raw/airports.csv") |>
  rename(airport_name = name)

airports_raw_filtered <- airports_raw[ ,c("faa","airport_name")] 

# joining data

flights_raw_filtered <- flights_raw[ ,c("flight","carrier","origin","dest","time_hour","dep_delay","arr_delay")] |>
  mutate(flight_id = row_number()) |>
  left_join(airlines_raw, by = "carrier") |>
  left_join(airports_raw_filtered, by = join_by("origin" == "faa")) |>
  rename(airport_origin_name = airport_name) |>
  left_join(airports_raw_filtered, by = join_by("dest" == "faa")) |>
  rename(airport_destination_name = airport_name)

```
### Delay classification

- Arrival delay > 60 minutes is significant
- Arrival delay 15 to 60 minutes is high
- Arrival delay 0 to 15 minutes is low
- Arrival delay -∞ to 0 is none 

```{r generating categorical variables, echo=FALSE}
# generating month, day, renaming 

flights_joined <- flights_raw_filtered[ , c("flight_id", "flight", "airline_name","airport_origin_name","airport_destination_name","dep_delay","arr_delay","time_hour")] |>
  mutate(day_of_week = wday(time_hour, label = TRUE, abbr = FALSE)) |>
  mutate(month = month(time_hour, label = TRUE, abbr = FALSE)) |>
  rename(date_time = time_hour) |>
  mutate(arrival_delay = case_when(
    arr_delay > 60 ~ "Significant",
    arr_delay > 15 ~ "High",
    arr_delay > 0 ~ "Low" , 
    TRUE ~ "None"
     )) |>
  mutate(departure_delay = case_when(
    dep_delay > 60 ~ "Significant",
    dep_delay > 15 ~ "High",
    dep_delay > 0 ~ "Low",
    TRUE ~ "None"
  ))
#case_when works top down
```


```{r duplicates and invalid flight removal, echo=FALSE}

# removing invalid flights and duplicates

flights_clean <- flights_joined |>
  filter(!is.na(dep_delay)) |>
  filter(!is.na(arr_delay)) |>
  filter(!is.na(airport_destination_name)) |>
  filter(!is.na(airport_origin_name)) |>
  filter(!is.na(airline_name)) |>
  distinct(airline_name, flight, airport_origin_name, airport_destination_name, dep_delay, arr_delay, date_time, .keep_all = TRUE) |>
  filter(year(date_time) == 2013)

summary(flights_clean)
glimpse(flights_clean)

# save data as parquet

write_parquet(flights_clean, "data/clean/flights_cleaned.parquet")
```

### Exploratory analysis

Flights were grouped by airline, departure airport, day of week, and departure hour. For each group, the percentage of flights experiencing high or significant arrival delays was calculated. The following analysis evaluates how delay risk varies across key booking decisions available to stakeholders.

```{r create functions for plot generation}
# read parquet

flights_cleaned <- read_parquet("data/clean/flights_cleaned.parquet")

glimpse(flights_cleaned)

# create function for plotting column charts for airport origin, airline and day of the week

plot_delay_risk <- function(data, group_var, plot_title, x_label) {
  data |>
    group_by({{ group_var }}, arrival_delay) |>
    summarise(n = n(), .groups = "drop") |>
    group_by({{ group_var }}) |>
    mutate(percent = n / sum(n) * 100) |>
    filter(arrival_delay %in% c("High", "Significant")) |>
    
    ggplot(aes(
      x = reorder({{ group_var }}, percent),
      y = percent,
      fill = arrival_delay
    )) +
    geom_col(position = "dodge") +
    coord_flip() +
    labs(
      x = x_label,
      y = "Percent of Flights",
      fill = "Delay Severity",
      title = plot_title
    )
  
}

# assign plot for delays by hour of day


delay_hour_pct_plot <- flights_cleaned |>
  mutate(hour_of_day = hour(date_time)) |>
  group_by(hour_of_day, arrival_delay) |>
  summarise(n = n(), .groups = "drop") |>
  group_by(hour_of_day) |>
  mutate(percent = (n/sum(n)*100)) |>
  filter(arrival_delay %in% c("High", "Significant")) |>
  
  ggplot(aes(
    x = hour_of_day,
    y = percent,
    color = arrival_delay
  )) +
  geom_line() +
  geom_point() +
  labs(
    x = "time",
    y = "Percent of Flights",
    color= "Delay Severity",
    title = "Risk Meaningful delays over time of day"
  ) 


```


Airline Delay risk

```{r airline delay plot, echo=FALSE}

plot_delay_risk(flights_cleaned, airline_name, "Risk of meaningful arrival delays by airlines", "Airline")

```

Insight
Substantial variation exists for both high and significant delay between the different airlines. This difference is clear when comparing Hawaiian Ailinces, Alaska Airlines and US airways to Frontier Airlines, Airtran Airways and Mesa Airlines.

Reccomendation
Stakeholders should prefer airlines with demonstratable lower delay risk when possible. 


Airport Origin Delay risk
```{r airport delay plot, echo=FALSE}
plot_delay_risk(flights_cleaned, airport_origin_name, "Risk of meaningful arrival delays by origin airport", "Airport Origin")
```

Insight
Small variation exists for both high and significant delay between the three airports. La Guardia demonstrates lower Significant and high delay severity. 

Reccomendation
Stakeholders should prefer La Guardia for their departing airport when possible. 


Day of the week delay plot

```{r day delay plot, echo=FALSE}

plot_delay_risk(flights_cleaned, day_of_week, "Risk of meaningful arrival delays by day", "Day")

```

Insight
Moderate delay variation exists between the days. Saturday and Sunday show the lowest delay risk while Thursday has he highest.

Reccomendation
Stakeholders should prioritize travelling on lower delay risk days. 

Hour of day delay plot

```{r hour delay plot, echo=FALSE}

print(delay_hour_pct_plot)

```
Insight
Substantial delay variation exists throughout the course of the day. Just before 10am experiences low delay which increases gradually until after midnight, which is when the highest delay risks are experienced.

Reccomendation
Stakeholders should prioritize travelling in the morning between 5am and 10am. 


### Key findings
- Early morning flights experience the lowest delays
- Flights incur gradually increasing delay risk throughout the day.
- Meaningful delay risk difference between the airlines
- Day of departure choice influences delay risk

### Recommendations
Stakeholders should primarily target time of day departure of around 5-10am. Of the flights available at that time, prioritize the lowest risk airlines, particularly US Airways, Alaska Airlines and Hawaiian Airlines. Next, prioritize the day of departure for Saturday or Sunday. Finally prioritize departing from La Guardia over the other two. 

### Limitations
Analysis limited to 2013 data. Results describe historical patterns, not predictions. 

### Future Work
Incorporate weather data. Model predictive delay probability. Extend analysis across multiple years. 
